Monoidal Toposes of Grammars and Semantic Actions

### 2022-08-15 notes.tex
description of inductive grammars.

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}

@@

This has a unit as well, which is the language of just the empty
string (with trivial parse tree/semantic action):
\[ I w = w = \epsilon \]

@@

Next, $\Lang$ has all limits and colimits. Coproducts give us the
``union'' of languages, with initial object the empty language, and
products give us the intersection, with the terminal object being the
language that accepts all strings, but with no information in the
parse.
\[ (g_1 + g_2) w = (g_1 w) + (g_2 w) \]
\[ 0 w = \emptyset \]
\[ (g_1 \times g_2) w =  g_1 w \times g_2 w\]
\[ 1 w =  1 \]

We can also define inductively defined grammars (and so access regular
and context free languages) as colimits, for instance the Kleene star
operation is the colimit of the $\omega$-diagram
\[ 0 \to (I + g * 0) \to I + g \sepconj (I + g \sepconj 0) \to \cdots \]

@@

By general properties of presheaf categories, the category of
languages has \emph{closed} structure with respect to the monoidal
products of sequencing $\sepconj$ and product $\times$.

First, the product has a right (multi-variable) adjoints:
\[ (L_1 \times L_2 \to  L_3) \cong L_1 \to (L_2 \Rightarrow L_3) \cong L_2 \to (L_1 \Rightarrow L_3) \]

Defined by
\[ (L \Rightarrow L')w = \textrm{Hom}((Y w \times L), L') = Lw \to L'w \]
Informally, $(L \Rightarrow L')w$ is constructive \emph{implication}
of languages: a parse of $w$ is a function that takes $L$-parses of
$w$ to $L'$-parses of $w$

Next, the sequencing operator $\sepconj$, since it is not symmetric, has
two right multi-variable adjoints:
\[ (L_1 \sepconj L_2 \to L_3) \cong (L_1 \to L_3 \lsepimp L_2) \cong (L_2 \to L_2 \rsepimp L_3) \]

Defined by
\[ (L \multimap L')w = \Pi_{w_l} L w_l \to L'(w_lw) \]
\[ (L' \leftmultimap L)w = \Pi_{w_r} L w_r \to L'(ww_r) \]

A parse of $(L \multimap L')w$ is a kind of ``partial'' parse of $L'$:
given any parse $L w_l$, we can get a complete parse $L' (w_lw)$. So
if $(L \multimap L') w$ successfully parses (and $L$ is non-empty),
then we have succesfully parsed a suffix of $L'$, that when
concatenated with a prefix that can by parsed as $L$ will yield a full
parse of $L'$.

Dually, the left hom $(L' \leftmultimap L)$ is a partial parse from
the right of the string.

These left and right hom operations can be used to define the
\emph{derivative} of a language with respect to a character or
word. If we focus on left-to-right parsing this would be:
\[ d^l_c(L) = Yc \multimap L \]
Why?
\[ (Yc \multimap L)w = \Pi_{w_l} \{ 1 | c = w_l \} \to L(w_lw) \cong L(cw) \]
I.e., a parse of $d^l_c(L)$ is precisely a parse of the string with
$c$ prepended.
%
This allows us, for any particular string $w$, to reduce the problem
of parsing $w$ with a grammar $g$ to the problem of producing a parse
of an empty string for the grammar $d^l_w(g)$. Then the typical
derivation of the Brzozowski derivative of a regular language follows
from abstract nonsense?

If we reduce to language recognition, this is a de Morgan-dual of the
``quotient'' of a formal language.
\[ (L_1 / L_2) w \iff \exists w' \in L_2. L_1(ww') \iff \neg (\forall w' \in L_2. \neg L_1(ww')) \iff \neg ((\neg L_1 \leftmultimap L_2) w) \]

%% A grammar $g$ is
%% \begin{enumerate}
%% \item \emph{total} if for every $w$ there is at least one parse
%% \item \emph{deterministic} if for every $w$ there is at most one parse
%% \item \emph{empty} if for every $w$ there are no parses
%% \end{enumerate}

%% %% \begin{definition}
%% %%   A grammar $g$ is \emph{uniquely decidable} if there exists a
%% %%   ``complement'' $g'$ such that $g \vee g'$ is total and
%% %%   deterministic.
%% %% \end{definition}

@@

>>>>

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

@@

This has a unit as well, which is the grammar with a unique parse of
the empty string (with trivial semantic action):
\[ I w = w = \epsilon \]

@@

Next, $\Grammar$ has all limits and colimits. Binary coproducts give
us the ``sum'' of grammars, i.e., a parse is a parse in one of the two
languages. The initial object is the empty grammar, the grammar that
has no parses.
\[ (g_1 + g_2) w = (g_1 w) + (g_2 w) \]
\[ 0 w = \emptyset \]
The semantic actions ``lie over'' the corresponding coproduct in
$\Set$. These correspond to the union and empty languages in $\Lang$.

Products give us a kind of intersection of grammars, where a parse
tree is a parse from each language, with the terminal object being
the language that accepts all strings with a unique parse.
\[ (g_1 \times g_2) w =  g_1 w \times g_2 w\]
\[ 1 w =  1 \]
These lie over the product and unit semantic actions.

Inductively defined grammars should be initial algebras of functors $F
: \Grammar \to \Grammar$. To define regular and context free grammars
it should suffice to use Ad\'amek's construction generalizing Kleene's
fixed point theorem to categories. That is, if $F$ is
$\omega$-co-continuous, i.e., preserves $\omega$-colimits, then it has
an initial algebra, which can be constructed as the colimit of
\[ F^0(0) \to F^1(0) \to F^2(0) \to \cdots \]

Then, for example the Kleene star of a grammar $g$ is given by the
initial algebra of the functor $F(\gamma) = I + g \sepconj \gamma$,
which is $\omega$-co-continuous because $\sepconj$ and $+$ are
$\omega$-co-continuous.

So far the interesting structure has mostly been in the category of
grammars, and the corresponding sets of parses have been isomorphic to
the set of parses. The simple combinator that we need is a ``map'' for
semantic actions, that says given a semantic action $A$ that produces
$X$ values, and a function $f : X \to Y$, we can push forward and get
an action $f_*A$ that produces $Y$ values. If $A = (a, g, X)$ then we
can define this simply as
\[ f_*(a,g,X) = (f \circ a, g, Y) \]

@@

By general properties of presheaf categories, the category of
grammars has \emph{closed} structure with respect to the monoidal
products of sequencing $\sepconj$ and product $\times$.

First, the product has a right (multi-variable) adjoint:
\[ (g_1 \times g_2 \to  g_3) \cong g_1 \to (g_2 \Rightarrow g_3) \cong g_2 \to (g_1 \Rightarrow g_3) \]

Defined by
\[ (g \Rightarrow g')w = \textrm{Hom}((Y w \times g), g') = gw \to g'w \]
Informally, $(g \Rightarrow g')w$ is constructive \emph{implication}
of languages: a parse of $w$ is a function that takes $g$-parses of
$w$ to $g'$-parses of $w$

Next, the sequencing operator $\sepconj$, since it is not symmetric, has
two right multi-variable adjoints:
\[ (g_1 \sepconj g_2 \to g_3) \cong (g_1 \to g_3 \lsepimp g_2) \cong (g_2 \to g_2 \rsepimp g_3) \]

Defined by
\[ (g \rsepimp g')w = \Pi_{w_l} g w_l \to g'(w_lw) \]
\[ (g' \lsepimp g)w = \Pi_{w_r} g w_r \to g'(ww_r) \]

A parse of $(g \rsepimp g')w$ is a kind of ``partial'' parse of $g'$:
given any parse $g w_l$, we can get a complete parse $g' (w_lw)$. So
if $(g \rsepimp g') w$ successfully parses (and $g$ is non-empty),
then we have succesfully parsed a suffix of $g'$, that when
concatenated with a prefix that can by parsed as $g$ will yield a full
parse of $g'$.
%
This Day convolution closed structure is used in separation logic.

Dually, the left hom $(g' \lsepimp g)$ is a partial parse from
the right of the string.

These left and right hom operations can be used to define the
\emph{derivative} of a language with respect to a character or
word. If we focus on left-to-right parsing this would be:
\[ d^l_c(g) = Yc \rsepimp g \]
Why?
\[ (Yc \rsepimp g)w = \Pi_{w_l} \{ 1 | c = w_l \} \to g(w_lw) \cong g(cw) \]
I.e., a parse of $d^l_c(g)$ is precisely a parse of the string with
$c$ prepended.
%
This allows us, for any particular string $w$, to reduce the problem
of parsing $w$ with a grammar $g$ to the problem of producing a parse
of an empty string for the grammar $d^l_w(g)$. Then the typical
derivation of the Brzozowski derivative of a regular language follows
from abstract nonsense?

If we reduce to language recognition, this is a de Morgan-dual of the
``quotient'' of a formal language.
\[ (g_1 / g_2) w \iff \exists w' \in g_2. g_1(ww') \iff \neg (\forall w' \in g_2. \neg g_1(ww')) \iff \neg ((\neg g_1 \lsepimp g_2) w) \]

\section{Application: Derivative of a Regular Semantic Action}

Let $A$ be a semantic action inductively built using the constructions
of separating conjunction, unit, binary and empty coproducts,
characters, Kleene star, and pushforward.

Claim:
\begin{theorem}
  If $A$ is a regular semantic action, then for any character $c$,
  \[ d_c(A) = Yc \rsepimp A \]
  is isomorphic to a regular semantic action.
\end{theorem}
\begin{proof}
  By induction on the definition of $A$.
  \begin{enumerate}
  \item $A = I$, then $Y c\rsepimp I \cong 0$
  \item $A = A_1 \sepconj A_2$, then
    \begin{align*}
      Yc \rsepimp A_1 \sepconj A_2
      \cong ((Yc \rsepimp A_1) \sepconj A_2) + ((A_1 \times I) \sepconj (Yc \rsepimp A_2))
    \end{align*}
  \item $A = f_*A'$, then
    \begin{align*}
      Y c \rsepimp f_*A \cong g_*(Yc \rsepimp A)
    \end{align*}
    Where $g : (1 \Rightarrow X) \to (1 \Rightarrow Y)$ is $g(h)(x) =
    f(h(x))$.
  \item TODO: finish
  \end{enumerate}
\end{proof}

@@

%% We can define some concepts on grammars using the internal language.

%% A grammar $g$ is
%% \begin{enumerate}
%% \item \emph{total} if for every $w$ there is at least one
%%   parse. Equivalently, $g$ has a global element.
%% \item \emph{deterministic} if for every $w$ there is at most one
%%   parse. In the internal language, $g$ is a proposition.
%% \item \emph{empty} if for every $w$ there are no parses. In the
%%   internal language, $g \cong 0$
%% \end{enumerate}

%% \begin{definition}
%%   A grammar $g$ is \emph{decidable} if its propositional truncation
%%   $||g||$ is a complemented truth value.
%% \end{definition}

%% \begin{definition}
%%   A grammar $g$ is \emph{uniquely decidable} if there exists a
%%   ``complement'' $\neg g$ such that $g \vee \neg g$ is total and
%%   deterministic.
%% \end{definition}
